AFQ-Insight

### 2019-12-13 reports/manuscript/abstract.tex
Respond to Mareike's edits, misc.

The white matter contains long-range connections between different
brain regions and the organization of these connections hold important
implications for brain function in health and disease. Tractometry
uses diffusion-weighted magnetic resonance imaging (dMRI) data
to quantify tissue properties (e.g. fractional anisotropy (FA),
mean diffusivity (MD), etc.), along the trajectories of these
connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the tract in each individual, or (b) performs analysis
point-by-point along each tract, with group comparisons or regression
models computed separately for each point along every one of the tracts.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the tracts, and selects informative
features by enforcing sparsity, not only at the level of individual
tracts, but also across the entire set of tracts and all of the measured
tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme that guards
against over-fitting and simultaneously identifies the correct
level of sparsity. We demonstrate the accuracy of the method in two
settings: in a classification setting, patients with amyotrophic
lateral sclerosis (ALS) are accurately distinguished from matched
controls\cite{sarica2017corticospinal} and SGL automatically identifies
FA in the corticospinal tract as important for this classification --
correctly finding the parts of the white matter known to be affected
by the disease. In a regression setting, dMRI is used to accurately
predict ``brain age'' \cite{yeatman2014lifespan, Brown2012-so}. In this
case, the weights are distributed throughout the white matter indicating
that many different regions of the white matter change with development
and contribute to the prediction of age. Thus, SGL makes it possible
to leverage the multivariate relationship between diffusion properties
measured along multiple tracts to make accurate predictions of subject
characteristics while simultaneously discovering the most relevant
features of the white matter for the characteristic of interest.

>>>>

The white matter contains long-range connections between different
brain regions and the organization of these connections hold important
implications for brain function in health and disease. Tractometry
uses diffusion-weighted magnetic resonance imaging (dMRI) data
to quantify tissue properties (e.g. fractional anisotropy (FA),
mean diffusivity (MD), etc.), along the trajectories of these
connections\cite{yeatman2012tract}. Statistical inference from
tractometry data usually either (a) averages these quantities along
the length of the tract in each individual, or (b) performs analysis
point-by-point along each tract, with group comparisons or regression
models computed separately for each point along every one of the tracts.
In the present work, we developed a method based on the sparse group
lasso (SGL) \cite{simon2013sparse} that takes into account tissue
properties measured along all of the tracts, and selects informative
features by enforcing sparsity, not only at the level of individual
tracts, but also across the entire set of tracts and all of the measured
tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme that guards
against over-fitting and simultaneously identifies the correct
level of sparsity. We demonstrate the accuracy of the method in two
settings: i) In a classification setting, patients with amyotrophic
lateral sclerosis (ALS) are accurately distinguished from matched
controls \cite{sarica2017corticospinal}. Furthermore, SGL automatically
identifies FA in the corticospinal tract as important for this
classification -- correctly finding the parts of the white matter known
to be affected by the disease. ii) In a regression setting, dMRI is
used to accurately predict ``brain age'' \cite{yeatman2014lifespan,
Brown2012-so}. In this case, the weights are distributed throughout the
white matter indicating that many different regions of the white matter
change with development and contribute to the prediction of age. Thus,
SGL makes it possible to leverage the multivariate relationship between
diffusion properties measured along multiple tracts to make accurate
predictions of subject characteristics while simultaneously discovering
the most relevant features of the white matter for the characteristic of
interest.

### 2019-12-13 reports/manuscript/sections/introduction.tex
Respond to Mareike's edits, misc.

Leveraging this fact, one of the most powerful methods currently
available to put macro- and micro-structure together is
\emph{tractometry}: assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are
several different available implementations of this overall idea,
the principles are similar \cite{yeatman2012tract, Yendiki2011-ay,
Wassermann2016-iv, ODonnell2009-uu}: tractometry begins by delineating
the parts of the white matter that belong to different major tracts or
axonal nerve fiber bundles, such as the corticospinal tract or arcuate
fasciculus, and samples the biophysical properties (such as fractional
anisotropy or mean diffusivity) along the length of these bundles.
In some cases, tissue properties along the length of each bundle are
summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability along
the trajectory of each bundle. This justifies retaining the individual
samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu}. However, while this retains important
information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals
to each other. This is usually done following one of the following
approaches:

>>>>

Leveraging this fact, one of the most powerful methods currently
available to put macro- and micro-structure together is
\emph{tractometry}: assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are
several different available implementations of this overall idea,
the principles are similar \cite{yeatman2012tract, Yendiki2011-ay,
Wassermann2016-iv, ODonnell2009-uu}: tractometry begins by delineating
the parts of the white matter that belong to different major tracts
(i.e. bundles of streamlines), such as the corticospinal tract or arcuate
fasciculus, and samples the biophysical properties (such as fractional
anisotropy or mean diffusivity) along the length of these bundles.
In some cases, tissue properties along the length of each bundle are
summarized by taking the mean along each bundle, but there is a large
body of evidence showing that there is systematic variability along
the trajectory of each bundle. This justifies retaining the individual
samples along the length of each bundle \cite{yeatman2012tract,
colby2012, ODonnell2009-uu}. However, while this retains important
information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals
to each other. This is usually done following one of the following
approaches:

### 2019-12-13 reports/manuscript/sections/methods.tex
One more pass through
Respond to Mareike's edits, misc.

\item Diffusion MRI data from a previous study of properties of the white matter
across the lifespan\cite{yeatman2014lifespan}, containing dMRI data from 76
subjects with ages 6-50. These data were measured in a GE Discovery 750 3T MRI
scanner at the Stanford Center for Cognitive and Neurobiological Imaging. The
Stanford University IRB approved the procedures of this study. Informed consent
was obtained from each adult participant, and assent for participation was
provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$mm^3$ with 96 non-colinear directions measured with a $b=2000$
$\frac{sec}{mm^2}$ and 30 non-colinear directions measured with a $b=1000$
$\frac{sec}{mm^2}$. Data was preprocessed to correct for subject motion and the
diffusion tensor model \cite{basser1994mr} was fit in every voxel, using a
robust fit \cite{chang2005restore}. These data were acquired using a dual spin
echo sequence, in which there is sufficient time for eddy currents to subside
between the application of the gradients and the image acquisition, so no eddy
current correction was appliced. We will refer to this dataset as WH.

\item Diffusion MRI from from a previous study of the corticospinal tract
profile in patients with amyotrophic lateral sclerosis
(ALS)\cite{sarica2017corticospinal}, containing data from 24 ALS patients and 24
demographically matched healthy controls. These data were measured in a GE
Discovery 750 3T MRI scanner at the Institute of Bioimaging and Molecular
Physiology in Catanzaro. Informed consent was provided as approved by the
Ethical Committee of the University ``Magna Graecia'' of Catanzaro. Voxel
resolution was \num{2x2x2} $mm^3$ and 27 non-colinear directions were
measured with a $b=1000$ $\frac{sec}{mm^2}$. Here, both motion correction and
eddy current correction were applied before modeling the diffusioon tensor in
each voxel. We will refer to this dataset as ALS.

@@

Data from both of these studies was processed in a similar manner, using the
Matlab Automated Fiber Quantification toolbox (AFQ) \cite{yeatman2012tract}:
streamlines representing fascicles of white matter tracts were generated
using a determinstic tractography algorithm that follows the prinicpal diffusion
direction of the diffusion tensor in each voxel (STT \cite{basser2000vivo}).
Major tracts were identified using multiple criteria: streamlines are selected
as candidates for inclusion in a bundle of streamlines that represents a tract
if they pass through known inclusion ROIs and do not pass through exclusion ROIs
\cite{wakana2007reproducibility}. In addition, a probabilistic atlas is used to
exclude streamlines which are unlikely to be part of a tract \cite{Hua2008-sh}.
Each streamline is resampled to 100 nodes and the robust mean at each location
is calculated by estimating the 3D covariance of the location of each node and
excluding streamlines that are more than 5 standard deviations from the mean
location in any node. Finally, a tract profile of tissue properties in each
tract was created by interpolating the value of MRI maps of these tissue
properties to the location of the nodes of the resampled streamlines designated
to each tract. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the mahalnobis
distance of the node from the average  of that node across streamlines. This
means that streamlines that are more representative of the tract contribute more
to the tract profile, relative to streamlines that are on the edge of the tract.

@@

Before fitting a model to the data, imputation and standardization are
performed. Missing node values (e.g., in cases where AFQ designates a node as
not-a-number) are imputed via linear interpolation. Care is taken not to
interpolate across the boundaries between different tracts. Some diffusion
metrics will have naturally larger variance than others and may therefore
dominate the objective function and make the SGL estimator unable to learn from
the lower variance metrics. For example, fractional anisotropy (FA) is bounded
between zero and one and could be overwhelmed by an unscaled higher-variance
metric like the mean diffusivity (MD). To prevent this we remove each feature's
mean and scale it to unit variance (z-score) using the
\lstinline{StandardScaler} from scikit-learn\cite{scikit-learn}. Scaling is
performed separately within each cross-validation set's training or testing data
to prevent leakage of information between the testing and training
sets\cite{kaufman2012leakage}.

@@

Often, the target variable $y$ is not in the domain in which the linear
model can be best fit to it. Equation \ref{eq:lm-approx} can be slightly
modified as:
\begin{equation}
    \hat{y} = f^-1(\mathbf{X} \hat{\beta}),
    \label{eq:lm-transform}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform
applied to the data before fitting the linear coefficients. For example,
an often-used transform is the logarithmic transform:
\begin{equation}
    f(\hat{y}) = \log_n(\hat{y})
    \label{eq:log-nonlinearity}
\end{equation}
In this case, the model is parametrized by one additional fit parameter,
$n$.

@@

For given values of $\lambda_1$ and $\lambda_2$, the cost function in equation
\ref{eq:sgl} can be optimized using proximal gradient descent methods
\cite{parikh2014proximal} here implemented as a custom proximal operator that is
then optimized using the C-OPT library\cite{copt}. This supplies an estimate of
the optimal $\hat{\beta}$ given a particular set of values for the
meta-parameters $\lambda_1$ and $\lambda_2$.

To objectively evaluate the model and guard against over-fitting, we used a
nested cross-validation scheme, depicted in Fig~\ref{fig:cross-val} for the
categorical classification case. The input data is shuffled and then decomposed
into $k$ batches, hereafter referred to as folds. For the ALS dataset we used
$k=10$ and for the WH dataset $k=5$. For each unique fold, we hold that fold out
as the test\textsubscript{outer} set and let the remaining data comprise the
train\textsubscript{outer} set, with the subscript indicating the depth of the
nested decomposition. We further decompose each train\textsubscript{outer} set
into three folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit an SGL
model using fixed regularization meta-parameters $\lambda_1$ and $\lambda_2$,
training the model using train\textsubscript{inner} and evaluating the fit on
test\textsubscript{inner}. We find the optimal values for $\lambda_1$ and
$\lambda_2$ using hyperoptimization, implemented using the hyperopt
library's \verb|fmin| function\cite{Bergstra_2015} with a tree of Parzen
estimators search algorithm\cite{bergstra2011algorithms}. For continuous
numerical $y$, \verb|fmin| searches for meta-parameter values that minimize the
median absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination ($R^2$).
For binary categorical $y$ \verb|fmin| seeks to maximize the classification
accuracy. This can also be done maximizing the area under the receiver operating
curve (ROC AUC) or the average precision. Using hyperoptimization, we find
optimal regularization parameters and $\hat{\beta}$ for each
train\textsubscript{outer} set and then use those to predict values for data in
test\textsubscript{outer}. Thus each subject in the dataset has a predicted
phenotype derived from a model that never saw its particular subject's data.

>>>>

\item Diffusion MRI from a previous study of the corticospinal
tract profile in patients with amyotrophic lateral sclerosis
(ALS \cite{sarica2017corticospinal}), containing data from 24 ALS
patients and 24 demographically matched healthy controls. These data
were measured in a GE Discovery 750 3T MRI scanner at the Institute
of Bioimaging and Molecular Physiology in Catanzaro. Informed consent
was provided as approved by the Ethical Committee of the University
``Magna Graecia'' of Catanzaro. Voxel resolution was \num{2x2x2}
$\text{mm}^3$ and 27 non-colinear directions were measured with a
$b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. Here, both motion correction
and eddy current correction were applied before modeling the diffusion
tensor in each voxel. We will refer to this dataset as ALS.

\item Diffusion MRI data from a previous study of properties of
the white matter across the lifespan \cite{yeatman2014lifespan},
containing dMRI data from 76 subjects with ages 6-50. These data were
measured in a GE Discovery 750 3T MRI scanner at the Stanford Center
for Cognitive and Neurobiological Imaging. The Stanford University
IRB approved the procedures of this study. Informed consent was
obtained from each adult participant, and assent for participation
was provided by parents/guardians for children. Voxel resolution was
\num{2x2x2}$\text{mm}^3$ with 96 non-colinear directions measured with a
$b=2000$ $\frac{\text{sec}}{\text{mm}^2}$ and 30 non-colinear directions
measured with a $b=1000$ $\frac{\text{sec}}{\text{mm}^2}$. Data was
preprocessed to correct for subject motion and the diffusion tensor
model \cite{basser1994mr} was fit in every voxel, using a robust fit
\cite{chang2005restore}. These data were acquired using a dual spin echo
sequence, in which there is sufficient time for eddy currents to subside
between the application of the gradients and the image acquisition, so
no eddy current correction was appliced. We will refer to this dataset
as WH.

@@

Data from both of these studies was processed in a similar manner,
using the Matlab Automated Fiber Quantification toolbox (AFQ)
\cite{yeatman2012tract}: streamlines representing fascicles of white
matter tracts were generated using a determinstic tractography algorithm
that follows the prinicpal diffusion direction of the diffusion tensor
in each voxel (STT) \cite{basser2000vivo}. Major tracts were identified
using multiple criteria: streamlines are selected as candidates for
inclusion in a bundle of streamlines that represents a tract if they
pass through known inclusion ROIs and do not pass through exclusion
ROIs \cite{wakana2007reproducibility}. In addition, a probabilistic
atlas is used to exclude streamlines which are unlikely to be part of
a tract \cite{Hua2008-sh}. Each streamline is resampled to 100 nodes
and the robust mean at each location is calculated by estimating the 3D
covariance of the location of each node and excluding streamlines that
are more than 5 standard deviations from the mean location in any node.
Finally, a tract profile of tissue properties in each tract was created
by interpolating the value of MRI maps of these tissue properties to the
location of the nodes of the resampled streamlines designated to each
tract. In each of 100 nodes, the values are summed across streamlines,
weighting the contribution of each streamline by the inverse of the
mahalnobis distance of the node from the average of that node across
streamlines. This means that streamlines that are more representative of
the tract contribute more to the tract profile, relative to streamlines
that are on the edge of the tract.

@@

Before fitting a model to the data, imputation and standardization are
performed. Missing node values (e.g., in cases where AFQ designates a node as
not-a-number) are imputed via linear interpolation. Care is taken not to
interpolate across the boundaries between different tracts. Some diffusion
metrics will have naturally larger variance than others and may therefore
dominate the objective function and make the SGL estimator unable to learn from
the lower variance metrics. For example, fractional anisotropy (FA) is bounded
between zero and one and could be overwhelmed by an unscaled higher-variance
metric like the mean diffusivity (MD). To prevent this we remove each feature's
mean and scale it to unit variance (z-score) using the
\lstinline{StandardScaler} from scikit-learn \cite{scikit-learn}. Scaling is
performed separately within each cross-validation set's training or testing data
to prevent leakage of information between the testing and training
sets\cite{kaufman2012leakage}.

@@

Often, the target variable $y$ is not in the domain in which the linear
model can be best fit to it. Equation \eqref{eq:lm-approx} can be slightly
modified as:
\begin{equation}
    \hat{y} = f^-1(\mathbf{X} \hat{\beta}),
    \label{eq:lm-transform}
\end{equation}
where the transformation function $f^{-1}$ characterizes the transform
applied to the data before fitting the linear coefficients. For example,
an often-used transform is the logarithmic transform:
\begin{equation}
    f(\hat{y}) = \log_n(\hat{y})
    \label{eq:log-nonlinearity}
\end{equation}
In this case, the model is parametrized by one additional fit parameter,
$n$.

@@

For given values of $\lambda_1$ and $\lambda_2$, the cost function in Eq~\eqref{eq:sgl} can be optimized using proximal gradient descent methods
\cite{parikh2014proximal} here implemented as a custom proximal operator that is
then optimized using the C-OPT library\cite{copt}. This supplies an estimate of
the optimal $\hat{\beta}$ given a particular set of values for the
meta-parameters $\lambda_1$ and $\lambda_2$.

To objectively evaluate the model and guard against over-fitting,
we used a nested cross-validation scheme, depicted in
Fig~\ref{fig:cross-val} for the categorical classification case.
The subjects (i.e. rows of the feature matrix $\mathbf{X}$ in
Fig~\ref{fig:group-structure} and Eq~\eqref{eq:lm}) are shuffled and
then decomposed into $k$ batches, hereafter referred to as folds. For
the ALS dataset we used $k=10$ and for the WH dataset $k=5$. For each
unique fold, we hold that fold out as the test\textsubscript{outer} set
and let the remaining data comprise the train\textsubscript{outer} set,
with the subscript indicating the depth of the nested decomposition.
We further decompose each train\textsubscript{outer} set into three
folds, and again for each unique fold, we hold out that fold as the
test\textsubscript{inner} set and let the remaining data comprise the
train\textsubscript{inner} set. At level-1 of the decomposition, we fit
an SGL model using fixed regularization meta-parameters $\lambda_1$
and $\lambda_2$, training the model using train\textsubscript{inner}
and evaluating the fit on test\textsubscript{inner}. We find
the optimal values for $\lambda_1$ and $\lambda_2$ using
hyperoptimization, implemented using the hyperopt library's \verb|fmin|
function\cite{Bergstra_2015} with a tree of Parzen estimators search
algorithm\cite{bergstra2011algorithms}. For continuous numerical $y$,
\verb|fmin| searches for meta-parameter values that minimize the median
absolute error. This can also be done minimizing the root of the mean
squared error (RMSE) or to maximizing the coefficient of determination
($R^2$). For binary categorical $y$ \verb|fmin| seeks to maximize the
classification accuracy. This can also be done maximizing the area
under the receiver operating curve (ROC AUC) or the average precision.
Using hyperoptimization, we find optimal regularization parameters and
$\hat{\beta}$ for each train\textsubscript{outer} set and then use those
to predict values for data in test\textsubscript{outer}. Thus each
subject in the dataset has a predicted phenotype derived from a model
that never saw its particular subject's data.
